The CrypTool Script: \\ Cryptography, Mathematics and More \\ \vspace{30pt} {\large Background reading\\%*[-8pt] for the free e-learning program CrypTool\\ (10th edition -- distributed with CrypTool version 1.4.30)} \vspace{60pt}

### 2010-01-14 script/de/chapters/primes.tex
Script[de+en]: Factorisation of RSA-768 added plus some minor things.

Um zu verstehen, warum die Anzahl der Primzahlen so schnell wächst, obwohl sich die
Intervallgrenzen jeweils nur um einen Exponentenwert von 1  unterscheiden, werfen wir einen
Blick auf beide Komponenten der Formel: (1) $x$ und (2) $\ln{x}$.

>>>>

Um zu verstehen, warum die Anzahl der Primzahlen so schnell wächst, obwohl sich die
Intervallgrenzen jeweils nur um einen Exponentenwert von 1  unterscheiden, werfen wir
einen Blick auf beide Komponenten der rechten Seite der Formel: $x$ und $\ln{x}$.

### 2010-01-14 script/en/chapters/numbertheory.tex
Script[de+en]: Factorisation of RSA-768 added plus some minor things.

\newpage
\hypertarget{Chapter_ElementaryNT}{}
\chapter{Introduction to Elementary Number Theory with Examples}
\label{Chapter_ElementaryNT}
(Bernhard Esslinger, July 2001; Updates: Dec. 2001, June 2002, May 2003,
 May 2005, March 2006, June 2007, July 2009) \\

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\hypertarget{SecurityRSA}{}
\section[Considerations regarding the security of the RSA algorithm]
{Considerations regarding the security of the RSA algorithm\footnotemark}
    \footnotetext{Major
    parts of chapters~\ref{complexity} and \ref{chptSecurityParam} follow
    the article ``Vorz\"uge und Grenzen des RSA-Verfahrens'' written by
    F.~Bourseau, D.~Fox and C.~Thiel \cite{nt:Bourseau2002}.}
    \label{SecurityRSA}

@@

% --------------------------------------------------------------------------
\vskip +40pt
\subsection{Security parameters because of new algorithms}
\label{chptSecurityParam}
\nopagebreak
%\vskip +15 pt
%\paragraph{Factorization algorithms\footnotemark}
\paragraph[Factorization algorithms]
          {Factorization algorithms\footnotemark}
\footnotetext{%
    \index{NT, Learning Tool for Number Theory}%
    \index{Educational tool NT}%
    With the educational tool for number theory {\bf NT} you can gather more
    experience with current factorization algorithms
    (see learning unit 5.1-5.5, pages 1-15/15).\\
    NT can be called in CrypTool\index{CrypTool} via the menu path
    {\bf Indiv. Procedures \textbackslash{} Number Theory Interactive
    \textbackslash{} Learning Tool for Number Theory}.
    See appendix~\ref{s:appendix-Learn-NT}.
}
\nopagebreak
The complexity is basically determined by the length $l$ of the module $n$.
Higher values for this major parameter are oriented at the possibilities of the
current algorithms for factorization:

\begin{itemize}
\item In 1994 a 129-digit RSA module (428 bit), published in 1977, was factorized
      by a distributed implementation of the Quadratic Sieve algorithm (QS),
      developed 1982 by Pomerance. This effort took 8 months.

      Please refer to:
      \begin{list}{}{\setlength{\topsep}{-7 pt}}
        \item[] C. Pomerance:  
                {\em The quadratic sieve factoring algorithm} 
                \cite{nt:Pomerance1984}.
      \end{list}
      \vskip +12pt

\item In 1999 a 155-digit module (512 bit) was factored with an implementation
      of the General Number Field Sieve algorithm (GNFS) 
      \index{General Number Field Sieve (GNFS)}, developed by Buhler, 
      Lenstra and Pomerance.
      The GNFS is more efficient than QS if $n$ is longer
      than about 116 decimal digits. This effort took 5 months.

      Please refer to:
      \begin{list}{}{\setlength{\topsep}{-7 pt}}
        \item[] J.P. Buhler, H.W. Lenstra, C. Pomerance:  
                {\em Factoring integers with the number field sieve} 
                \cite{nt:Buhler1993}.
      \end{list}
      \vskip +12pt
\end{itemize}

This made evident that a module length of 512 bit no longer prevents from
attackers.

And also past 1999 further factorization progress was made up to now 
(see RSA-200 in chapter~\ref{RSA-200}).

@@

The module length is not the only parameter relevant for security.
Beneath requirements from implementation and engineering the sizes and
the proportions of the  parameters e, d and N are relevant.

According attacks based on lattice reductions\index{Lattice reduction} 
are a real threat for (too) simple implementations of RSA.
Theses attacks can be structured into the following four categories:
\begin{itemize}
   \item Attacks against very small public keys e (e.g.\ $e = 3$). 
   \item Attacks against relatively small private exponents d(e.g.\ $d < N ^ {0.5}$).
   \item Factorization of the modulus N, if one of the factors p or q is {\em partly} known.
   \item Attacks requiring, that a {\em part} of the private key d is known.
\end{itemize}

A good overview about the current status of publications concerning these attacks
can be found in the diploma thesis of Matthias Schneider \cite{nt:SchneiderM2004}.

@@

Within the last 20 years a lot of progress has been made. Estimations about
the future development of the ability to factor RSA modules vary and depend
on some assumptions:
\begin{itemize}
  \item progression in computing performance\index{Performance} (Moore's 
      law\index{Moore's law}\index{Moore, Gordon E.}: every 18 month the
      computing power will double) and in grid computing\index{Grid computing}.
  \item development of new algorithms.
\end{itemize}
Within the last years the module bit length feasible for factorization
increased --- even without new algorithms --- by 10 bit per year. Larger
numbers require not only more time to be factored, but also huge RAM
storage for the solutions matrix being used by the best algorithms known
today.  This need for storage grows like the square root of the computation
time, i.\:e.\ also sub-exponentially. Because RAM availability increased
exponentially in the recent decades, it seems that this should not be the
limiting factor.

@@

This forecast by Dirk Fox\index{Fox, Dirk} from the year 2001 seems to prove true
by the factorization record of RSA-200 (see chapter~\ref{RSA-200}). His estimation for the year 2005,
to achieve a bit length of 660 bit, was almost a precision landing  (compare 
figure~\ref{secorvo-factorization-forecast}). \\
If the forecast withstand in the future then the factorization of an RSA modulus
of 1024 bit can be expected in 15 years.

\begin{figure}[!hb]
\begin{center}
\frame{
\includegraphics[scale=0.64, clip, viewport=-20 0 680 520]{figures/PrognoseRSAFaktorisierungSecorvo.pdf}
}
\caption{Forecast about future factorization records compared with current results
(from Secorvo)}
\label{secorvo-factorization-forecast}
\end{center}
\end{figure}

@@

An exhaustive overview about the factoring records
\index{Factorization!factoring records} of composed integers using different
methods can be found on the following web pages:
\vspace{-10pt}
\begin{itemize}
\item[]
     \url{http://www.crypto-world.com} \\
     \url{http://www.tutorgig.com/ed/RSA\_number}  ~~ The RSA Factoring Challenge \\
     \url{http://en.wikipedia.org/wiki/Integer_factorization_records}
\end{itemize}
%\footnote{%
%This site was not quite up-to-date end of May 2005: RSA-200 was missing.}

The current record (as of May 2005) obtained using the GNFS method
(General Number Field Sieve) factorized a general 200 decimal digit into its
both prime factors.

\vskip +10pt
The last records\footnote{%
The 'RSA numbers' are certain large semiprime numbers (i.e., numbers with
exactly two prime factors)\index{Number!semi prime}. 
They were generated and published by the 
company RSA Security and they form the basis of the RSA Factoring 
Challenge, in which factorizations for these numbers are sought.\\
See \url{http://www.rsa.com/rsalabs/node.asp?id=2092}.

RSA Labs offers its challenges since the beginning of the 1990th.
The first RSA Factoring Challenge labeled the numbers, from RSA-100 
to RSA-500, according to their number of decimal digits; the second 
RSA Factoring Challenge labeled the numbers according to their number 
of binary digits. Within the second challenge cash prizes were 
offered for successful factorizations of RSA576 to RSA2048 (RSA576, 
RSA640 etc. using 64 bit steps upwards --- An exception to this is
RSA617, which was created prior to the change in the numbering scheme).\\
But the RSA challenges ended ahead of time in 2007, RSA Inc. retracted
the prize.
  \index{Challenge}
  \index{Cipher challenge}
  \index{Crypto challenge}

The 'C numbers' originate from the Cunningham project:
\url{http://www.cerias.purdue.edu/homes/ssw/cun/}.
                          }
with factorization algorithms for composed numbers are
listed in the table~\ref{factorizationrecords}.

\begin{table}[ht]
\begin{center}
\begin{tabular}{|c|cccc|}
\hline
	& {\bf Decimal digits} & {\bf Binary digits} & {\bf Factored on} & {\bf Factored by} \\
\hline
	C307 & 		307 & 1017 & May, 2007 & Jens Franke et al. \\
	RSA-200 & 	200 & 663 & May, 2005 & Jens Franke et al. \\
	RSA640\footnotemark & 	193 & 600 & November, 2005 & Jens Franke et al. \\
	C176 & 		176 & 583 & May, 2005 & Kazumaro Aoki et al. \\
	RSA576 & 	174 & 576 & December, 2003 & Jens Franke et al. \\
	RSA-160 & 	160 & 530 & April, 2003 & Jens Franke et al. \\
	C158 & 		158 & 523 & January, 2002 & Jens Franke et al. \\
	RSA-155	&	155 & 512 & August, 1999 & Herman te Riele et al. \\
\hline
\end{tabular}
\caption{The current factoring records (as of July 2009)}    % Eyecatcher_neue-Mersenne 
\label{factorizationrecords}
\end{center}
\end{table} 
\footnotetext{%
A research group of the GISA solved this challenge which was awarded with
20,000 US dollar using the GNFS method. The researchers needed about five
months to divide this number into its both 320 bit long prime factors.

@@

The next bigger challenge is RSA-704.
See \url{http://www.heise.de/newsticker/meldung/print/65957}.
}

@@

Below the last records are explained in more detail.
The two methods, GNFS and SNFS, used to do so are shortly illustrated at the 
following web pages:
\index{General Number Field Sieve (GNFS)}
\index{Special Number Field Sieve (SNFS)}
\vspace{-10pt}
\begin{itemize}
\item[]
      \href{http://en.wikipedia.org/wiki/Special_number_field_sieve}
   {\texttt{http://en.wikipedia.org/wiki/Special\_number\_field\_sieve}} \\
      \href{http://en.wikipedia.org/wiki/General_number_field_sieve}
   {\texttt{http://en.wikipedia.org/wiki/General\_number\_field\_sieve}}
\end{itemize}
%\vspace{-10pt}

@@

On May 9, 2005 the research group of Prof. Jens Franke at the German
University of Bonn\footnote{%
   \href{http://www.loria.fr/~zimmerma/records/rsa200}
{\texttt{http://www.loria.fr/\~{}zimmerma/records/rsa200}}} announced,
that they achieved a new world record in number factorization
together with their colleagues of the Amsterdam Centrum voor Wiskunde
en Informatica.

They factorized a 200-digit number into its both prime factors 
(these are build with each 100 decimal digits) using the GNFS method 
(General Number Field Sieve)\index{General Number Field Sieve (GNFS)}.

@@

@@

% be_2005_UPDATEN_if-new-mersenne-prime-appears % Eyecatcher_neue-Mersenne
Length in bits of the current world records:\\
% (RSA-200 $\longleftrightarrow{}$ 43rd known Mersenne prime) \\
%  ~  $663 \longleftrightarrow 30,402,457$.
%$$ (RSA-200 ~~ \longleftrightarrow{} ~~ 43rd ~known~Mersenne~Prime) $$ mnmnmn
%$$ (RSA~-200 ~~ \longleftrightarrow{} ~~ 43rd ~known~Mersenne~Prime) $$
$$ [RSA{-}200~Number] ~~\longleftrightarrow{}~~ [43rd ~known~Mersenne~Prime] $$
$$ 663 ~~ \longleftrightarrow{} ~~ 30,402,457 ~~~~~$$

@@

\bibitem[Agrawal2002]{nt:Agrawal2002}  \index{Agrawal 2002} 
    M. Agrawal, N. Kayal, N. Saxena, \\
    {\em PRIMES in P}, August 2002, revised paper: \\
       \href{http://www.cse.iitk.ac.in/~manindra/algebra/primality_v6.pdf}
    {\texttt{http://www.cse.iitk.ac.in/~manindra/algebra/primality\_v6.pdf}} \\
    See also the website "The AKS "PRIMES in P" Algorithm Resource":
       \href{http://fatphil.org/maths/AKS/}
    {\texttt{http://fatphil.org/maths/AKS/}}

@@

\bibitem[Bernstein2001]{nt:Bernstein2001} \index{Bernstein 2001}
    D.~J. Bernstein, \\
    {\em Circuits for integer factorization: a proposal},\\ 
    \href{http://cr.yp.to/papers/nfscircuit.ps}
    {\texttt{http://cr.yp.to/papers/nfscircuit.ps}} \\
    \href{http://cr.yp.to/djb.html}{\texttt{http://cr.yp.to/djb.html}}

@@

\bibitem[Bourseau2002]{nt:Bourseau2002} \index{Bourseau 2002} \index{Fox 2002}
    F. Bourseau, D. Fox, C. Thiel, \\
    {\em Vorz\"uge und Grenzen des RSA-Verfahrens},\\
    In: Datenschutz und Datensicherheit (DuD) 26/2002, pp~84-89 (see www.dud.de),
    \href{http://http://www.secorvo.de/publikationen/rsa-grenzen-fox-2002.pdf}
         {\texttt{http://www.secorvo.de/publikationen/rsa-grenzen-fox-2002.pdf}}

@@

\bibitem[GISA2002]{nt:GISA2002} \index{GISA 2002}
    GISA (German Information Security Agency), \\
    {\em Recommendation for key length selection}, Bonn, Sep. 2002, \\
    \href{http://www.bsi.bund.de/esig/basics/techbas/krypto/bund02v7.pdf}
    {\tt http://www.bsi.bund.de/esig/basics/techbas/krypto/bund02v7.pdf} \\
    A statement on these recommendations: \\
    \href{http://www.secorvo.de/publikat/stellungnahme-algorithmenempfehlung-020307.pdf}
    {\texttt{http://www.secorvo.de/publikat/}}\\
    \hspace*{1cm}
    \href{http://www.secorvo.de/publikat/stellungnahme-algorithmenempfehlung-020307.pdf}
    {\texttt{stellungnahme-algorithmenempfehlung-020307.pdf}}

@@

\bibitem[Lenstra1999]{nt:Lenstra1999} Arjen K. Lenstra, Eric R. Verheul
     \index{Lenstra/Verheul 1999} \\
     {\em Selecting Cryptographic Key Sizes (1999)},\\
     Journal of Cryptology: the journal of the International 
     Association for Cryptologic Research \\
     \href{http://www.cryptosavvy.com/cryptosizes.pdf}
     {\texttt{http://www.cryptosavvy.com/cryptosizes.pdf}}

\bibitem[Lenstra2002]{nt:Lenstra2002} \index{Lenstra 2002}
    Arjen K. Lenstra, Adi Shamir, Jim Tomlinson, Eran Tromer,\\
    {\em Analysis of Bernstein's Factorization Circuit},\\
    \href{http://www.cryptosavvy.com/mesh.pdf}
    {\texttt{http://www.cryptosavvy.com/mesh.pdf}}

@@

\bibitem[RSA Security 2002]{nt:RSA Security 2002} \index{RSA Security 2002} 
    RSA Security, \\
    {\em Has the RSA algorithm been compromised as a result 
    of Bernstein's Paper?}, \\
    April 8th, 2002, \\
    \href{http://www.rsasecurity.com/}{\tt http://www.rsasecurity.com/}

@@

\bibitem[Shamir2003]{nt:Shamir2003} \index{Shamir 2003} \index{TWIRL device} 
    Adi Shamir, Eran Tromer, \\
    {\em Factoring Large Numbers with the TWIRL Device}, 
    Januar 2003, \\
    \href{http://www.wisdom.weizmann.ac.il/~tromer/}
    {\texttt{http://www.wisdom.weizmann.ac.il/\~{}tromer/}}.

\bibitem[Shamir2003a]{nt:Shamir2003a} \index{Shamir 2003a} \index{TWIRL device} 
    Adi Shamir, Eran Tromer, \\
    {\em On the Cost of Factoring RSA-1024}, 
    RSA Laboratories CryptoBytes Volume 6, No. 2, Summer 2003, p. 11-20 \\
    \url{http://www.rsasecurity.com/rsalabs/cryptobytes/CryptoBytes_August_2003.pdf}

>>>>

\newpage
\hypertarget{Chapter_ElementaryNT}{}
\chapter{Introduction to Elementary Number Theory with Examples}
\label{Chapter_ElementaryNT}
(Bernhard Esslinger, July 2001; Updates: Dec. 2001, June 2002, May 2003,
 May 2005, March 2006, June 2007, July 2009, Jan. 2010) \\

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\newpage
\hypertarget{SecurityRSA}{}
\section[Considerations regarding the security of the RSA algorithm]
{Considerations regarding the security of the RSA algorithm\footnotemark}
    \footnotetext{%
    Major parts of chapter~\ref{SecurityRSA} follow
    the article ``Vorz\"uge und Grenzen des RSA-Verfahrens'' written by
    F.~Bourseau, D.~Fox and C.~Thiel \cite{nt:Bourseau2002}.}
    \label{SecurityRSA}

@@

% --------------------------------------------------------------------------
\vskip +40pt
\subsection{Security parameters because of new algorithms}
\label{chptSecurityParam}
\nopagebreak
%\vskip +15 pt
%\paragraph{Factorization algorithms\footnotemark}
\paragraph[Factorization algorithms]
          {Factorization algorithms\footnotemark}
\footnotetext{%
    \index{NT, Learning Tool for Number Theory}%
    \index{Educational tool NT}%
    With the educational tool for number theory {\bf NT} you can gather more
    experience with current factorization algorithms
    (see learning unit 5.1-5.5, pages 1-15/15).\\
    NT can be called in CrypTool\index{CrypTool} via the menu path
    {\bf Indiv. Procedures \textbackslash{} Number Theory Interactive
    \textbackslash{} Learning Tool for Number Theory}.
    See appendix~\ref{s:appendix-Learn-NT}.
}
\nopagebreak
The complexity is basically determined by the length $l$ of the modulus $n$.
Higher values for this major parameter are oriented at the possibilities of the
current algorithms for factorization:

\begin{itemize}
\item In 1994 a 129-digit RSA modulus (428 bit), published in 1977, was factorized
      by a distributed implementation of the Quadratic Sieve algorithm (QS),
      developed 1982 by Pomerance. This effort took 8 months.
      \index{Quadratic Sieve-Algorithmus (QS)}\\
      Please refer to:
      \begin{list}{}{\setlength{\topsep}{-7 pt}}
        \item[] C. Pomerance:  
                {\em The quadratic sieve factoring algorithm} 
                \cite{nt:Pomerance1984}.
      \end{list}
      \vskip +12pt

\item In 1999 a 155-digit modulus (512 bit) was factored with an implementation
      of the General Number Field Sieve algorithm (GNFS), 
      developed by Buhler, Lenstra and Pomerance.
      The GNFS is more efficient than QS if $n$ is longer
      than about 116 decimal digits. This effort took 5 months.
      \index{General Number Field Sieve (GNFS)}\\
      Please refer to:
      \begin{list}{}{\setlength{\topsep}{-7 pt}}
        \item[] J.P. Buhler, H.W. Lenstra, C. Pomerance:  
                {\em Factoring integers with the number field sieve} 
                \cite{nt:Buhler1993}.
      \end{list}
      \vskip +12pt

\item Ten years later, end of 2009, a 232-digit modulus (768 bit) was
      factored by Kleinjung etc. after 2 1/2 years.\\
       Please refer to:
      \begin{list}{}{\setlength{\topsep}{-7 pt}}
        \item[] T. Kleinjung, et. al.:  
                {\em Factorization of a 768-bit RSA modulus} 
                \cite{nt:Kleinjung2010}.
      \end{list}
      \vskip +12pt
\end{itemize}

This made practically evident that a module length of $768$ bit no longer
prevents from attackers.

Details about factorization progress since 1999 see chapter~\ref{RSA-200}.

@@

The module length is not the only parameter relevant for security.
Beneath requirements from implementation and engineering the sizes and
the proportions of the  parameters e, d and n are relevant.

According attacks based on lattice reductions\index{Lattice reduction} 
are a real threat for (too) simple implementations of RSA.
Theses attacks can be structured into the following four categories:
\begin{itemize}
   \item Attacks against very small public keys e (e.g.\ $e = 3$). 
   \item Attacks against relatively small private exponents d (e.g.\ $d < n ^ {0.5}$).
   \item Factorization of the modulus n, if one of the factors p or q is {\em partly} known.
   \item Attacks requiring, that a {\em part} of the private key d is known.
\end{itemize}

A good overview concerning these attacks can be found in the diploma thesis
of Matthias Schneider \cite{nt:SchneiderM2004}.

@@

Since 1980 a lot of progress has been made. Estimations about
the future development of the ability to factor RSA modules vary and depend
on some assumptions:
\begin{itemize}
  \item progression in computing performance\index{Performance} (Moore's 
      law\index{Moore's law}\index{Moore, Gordon E.}: every 18 month the
      computing power will double) and in grid computing\index{Grid computing}.
  \item development of new algorithms.
\end{itemize}
Within the last years the module bit length feasible for factorization
increased --- even without new algorithms --- by 10 bit per year. Larger
numbers require not only more time to be factored, but also huge RAM
storage for the solutions matrix being used by the best algorithms known
today.  This need for storage grows like the square root of the computation
time, i.\:e.\ also sub-exponentially. Because RAM availability increased
exponentially in the recent decades, it seems that this should not be the
limiting factor.

@@

This forecast by Dirk Fox\index{Fox, Dirk} from the year 2001 seems to prove true
by the latest factorization records of RSA-200 and RSA-768 (see chapter~\ref{RSA-200}).
His estimation for the year 2005,
to achieve a bit length of 660 bit, was almost a precision landing  (compare 
figure~\ref{secorvo-factorization-forecast}).

If the forecast withstands in the future then the factorization of an RSA modulus
of 1024 bit can be expected in the year 2020.

\begin{figure}[!hb]
\begin{center}
\frame{
\includegraphics[scale=0.64, clip, viewport=-20 0 680 520]
                 {figures/PrognoseRSAFaktorisierungSecorvo.png}
 }
\caption{Comparison of the published factorization records (blue) and of the
predicted development (red; by Fox 2001)}
\label{secorvo-factorization-forecast}
\end{center}
\end{figure}

@@

An exhaustive overview about the factoring records
\index{Factorization!factoring records} of composed integers using different
methods can be found on the following web pages:
\vspace{-10pt}
\begin{itemize}
\item[]
     \url{http://www.crypto-world.com} \\
     \url{http://www.tutorgig.com/ed/RSA\_number}  ~~ The RSA Factoring Challenge \\
     \url{http://en.wikipedia.org/wiki/Integer_factorization_records}\\
     \url{http://en.wikipedia.org/wiki/RSA_Factoring_Challenge}
\end{itemize}
%\footnote{%
%This site was not quite up-to-date end of May 2005: RSA-200 was missing.}

The current record (as of Jan. 2010) obtained using the GNFS method
(General Number Field Sieve) factorized a general 232 decimal digit into its
both prime factors.

\vskip +10pt
The last records\footnote{%
The 'RSA numbers' are certain large semiprime numbers (i.e., numbers with
exactly two prime factors)\index{Number!semi prime}. 
They were generated and published by the company RSA Security: In the
RSA Factoring Challenge the prime factors for these numbers are sought.\\
See \url{http://www.rsa.com/rsalabs/node.asp?id=2092}.

RSA Labs offers its challenges since the beginning of the 1990th.
The first RSA Factoring Challenge labeled the numbers, from RSA-100 
to RSA-500, according to their number of decimal digits; the second 
RSA Factoring Challenge labeled the numbers according to their number 
of binary digits. Within the second challenge cash prizes were 
offered for successful factorizations of RSA-576 to RSA-2048 (RSA-576, 
RSA-640 etc. using 64 bit steps upwards --- An exception to this is
RSA-617, which was created prior to the change in the numbering scheme).
But the RSA challenges ended ahead of time in 2007, RSA Inc. retracted
the prize.
  \index{Challenge}
  \index{Cipher challenge}
  \index{Crypto challenge}

The 'C numbers' originate from the Cunningham project:
\url{http://www.cerias.purdue.edu/homes/ssw/cun/}.
These are factors of Mersenne numbers, which have a very special
form. This  makes it an order of magnitude easier to factor them
as moduli of the same length build for RSA.
                          }
with factorization algorithms for composed numbers are
listed in the table~\ref{factorizationrecords}.

\begin{table}[ht]
\begin{center}
\begin{tabular}{|c|cccc|}
\hline
	& {\bf Decimal digits} & {\bf Binary digits} & {\bf Factored on} & {\bf Factored by} \\
\hline
	&&&& \\
	RSA-768 & 	232 & 768 & December, 2010 & Thorsten Kleinjung et al. \\
	RSA-200 & 	200 & 663 & May, 2005 & Jens Franke et al. \\
	RSA-640\footnotemark & 	193 & 640 & November, 2005 & Jens Franke et al. \\
	RSA-576 & 	174 & 576 & December, 2003 & Jens Franke et al. \\
	RSA-160 & 	160 & 530 & April, 2003 & Jens Franke et al. \\
	RSA-155	&	155 & 512 & August, 1999 & Herman te Riele et al. \\
	.........&&&& \\
	C307 & 		307 & 1017 & May, 2007 & Jens Franke et al. \\
	C176 & 		176 & 583 & May, 2005 & Kazumaro Aoki et al. \\
	C158 & 		158 & 523 & January, 2002 & Jens Franke et al. \\
\hline
\end{tabular}
\caption{The current factoring records (as of Jan. 2010)}    % Eyecatcher_neue-Mersenne 
\label{factorizationrecords}
\end{center}
\end{table} 
\footnotetext{%
A research group of the GISA solved this challenge which was awarded with
20,000 US dollar using the GNFS method. The researchers needed about five
months to divide this number into its both 320 bit long prime factors.

@@

See \url{http://www.heise.de/newsticker/meldung/print/65957}.
}

@@

\noindent Below the last records are explained in more detail.
The two methods, GNFS and SNFS, used to do so are shortly illustrated at the 
following web pages:
\index{General Number Field Sieve (GNFS)}
\index{Special Number Field Sieve (SNFS)}
\vspace{-10pt}
\begin{itemize}
\item[]
      \href{http://en.wikipedia.org/wiki/Special_number_field_sieve}
   {\texttt{http://en.wikipedia.org/wiki/Special\_number\_field\_sieve}} \\
      \href{http://en.wikipedia.org/wiki/General_number_field_sieve}
   {\texttt{http://en.wikipedia.org/wiki/General\_number\_field\_sieve}}
\end{itemize}
%\vspace{-10pt}

@@

On May 9, 2005 the research group of Prof. Jens Franke at the German
University of Bonn\footnote{%
   \url{http://www.loria.fr/~zimmerma/records/rsa200}} announced,
that they achieved to factorize a 200-digit number into its both prime factors 
(these are build with each 100 decimal digits) using the GNFS method 
(General Number Field Sieve)\index{General Number Field Sieve (GNFS)}.

@@

% --------------------------------------------------------------------------
\vskip +20pt
\hypertarget{RSA-768-chap3}{}
\paragraph{RSA-768} \label{RSA-768} \index{RSA-768}\mbox{}

On December 12, 2009 the research group around Prof. Thorsten
Kleinjung\footnote{%
   \url{http://eprint.iacr.org/2010/006.pdf} \cite{nt:Kleinjung2010}
         } announced,
that they achieved to factorize a 232-digit number into its both prime factors 
(both factors have 116 decimal digits). They used the GNFS method 
(General Number Field Sieve)\index{General Number Field Sieve (GNFS)} in a
way where they did ``oversieving'' on several hundred computers
before starting the matrix step.

The composite number called ``RSA-768'' is (the following 3 lines contain
one number):
$$
\begin{array}{c}
123018668453011775513049495838496272077285356959533479219732245215172640050726\\
365751874520219978646938995647494277406384592519255732630345373154826850791702\\
6122142913461670429214311602221240479274737794080665351419597459856902143413
\end{array}
$$
The factorization of RSA-768 resulted in the following two prime factors (each with 384 bit):
$$
\begin{array}{c}
p = 3347807169895689878604416984821269081770479498371376856891\\
    2431388982883793878002287614711652531743087737814467999489
\end{array}
$$
and
$$
\begin{array}{c}
q = 3674604366679959042824463379962795263227915816434308764267\\
    6032283815739666511279233373417143396810270092798736308917
\end{array}
$$

The calculations took about 2 1/2 years\footnote{%
This was an ``academic effort'' -- organisations with bigger resources
could do it much faster.}.
\vspace{24pt}

@@

% be_2005_UPDATEN_if-new-mersenne-prime-appears % Eyecatcher_neue-Mersenne
Length in bits of the current world records:\\
% (RSA-200 $\longleftrightarrow{}$ 43rd known Mersenne prime) \\
%  ~  $663 \longleftrightarrow 30,402,457$.
%$$ (RSA-200 ~~ \longleftrightarrow{} ~~ 43rd ~known~Mersenne~Prime) $$ mnmnmn
%$$ (RSA~-200 ~~ \longleftrightarrow{} ~~ 43rd ~known~Mersenne~Prime) $$
$$ [RSA{-}768~Number] ~~\longleftrightarrow{}~~ [47th ~known~Mersenne~Prime] $$
$$ 768 ~~ \longleftrightarrow{} ~~ 43,112,609 ~~~~~$$

@@

\bibitem[Agrawal2002]{nt:Agrawal2002}  \index{Agrawal 2002} 
    M. Agrawal, N. Kayal, N. Saxena, \\
    {\em PRIMES in P}, August 2002, revised paper: \\
       \url{http://www.cse.iitk.ac.in/~manindra/algebra/primality_v6.pdf}\\
    See also the website "The AKS "PRIMES in P" Algorithm Resource":
       \url{http://fatphil.org/maths/AKS/}.

@@

\bibitem[Bernstein2001]{nt:Bernstein2001} \index{Bernstein 2001}
    D.~J. Bernstein, \\
    {\em Circuits for integer factorization: a proposal},\\ 
    \url{http://cr.yp.to/papers/nfscircuit.ps} \\
    \url{http://cr.yp.to/djb.html}.

@@

\bibitem[Bourseau2002]{nt:Bourseau2002} \index{Bourseau 2002} \index{Fox 2002}
    F. Bourseau, D. Fox, C. Thiel, \\
    {\em Vorz\"uge und Grenzen des RSA-Verfahrens},\\
    In: Datenschutz und Datensicherheit (DuD) 26/2002, pp~84-89 (see www.dud.de),
    \url{http://www.secorvo.de/publikationen/rsa-grenzen-fox-2002.pdf}.

@@

\bibitem[GISA2002]{nt:GISA2002} \index{GISA 2002}
    GISA (German Information Security Agency), \\
    {\em Recommendation for key length selection}, Bonn, Sep. 2002, \\
    \url{http://www.bsi.bund.de/esig/basics/techbas/krypto/bund02v7.pdf} \\
    A statement on these recommendations: \\
    % \hspace*{1cm}
    \url{http://www.secorvo.de/publikat/stellungnahme-algorithmenempfehlung-020307.pdf}.

@@

\bibitem[Lenstra1999]{nt:Lenstra1999} Arjen K. Lenstra, Eric R. Verheul
     \index{Lenstra/Verheul 1999} \\
     {\em Selecting Cryptographic Key Sizes (1999)},\\
     Journal of Cryptology: the journal of the International 
     Association for Cryptologic Research, \\
     \url{http://www.cryptosavvy.com/cryptosizes.pdf}.

\bibitem[Lenstra2002]{nt:Lenstra2002} \index{Lenstra 2002}
    Arjen K. Lenstra, Adi Shamir, Jim Tomlinson, Eran Tromer,\\
    {\em Analysis of Bernstein's Factorization Circuit},\\
    \url{http://www.cryptosavvy.com/mesh.pdf}.

@@

\bibitem[RSA Security 2002]{nt:RSA Security 2002} \index{RSA Security 2002} 
    RSA Security, \\
    {\em Has the RSA algorithm been compromised as a result 
    of Bernstein's Paper?}, \\
    April 8th, 2002, \\
    \url{http://www.rsasecurity.com/}.

@@

\bibitem[Shamir2003]{nt:Shamir2003} \index{Shamir 2003} \index{TWIRL device} 
    Adi Shamir, Eran Tromer, \\
    {\em Factoring Large Numbers with the TWIRL Device}, 
    Januar 2003, \\
    \url{http://www.wisdom.weizmann.ac.il/~tromer/}.

\bibitem[Shamir2003a]{nt:Shamir2003a} \index{Shamir 2003a} \index{TWIRL device} 
    Adi Shamir, Eran Tromer, \\
    {\em On the Cost of Factoring RSA-1024}, 
    RSA Laboratories CryptoBytes Volume 6, No. 2, Summer 2003, p. 11-20, \\
    \url{http://www.rsasecurity.com/rsalabs/cryptobytes/CryptoBytes_August_2003.pdf}

### 2010-01-14 script/en/chapters/primes.tex
Script[de+en]: Factorisation of RSA-768 added plus some minor things.

To understand, why the number of primes is growing so rapidly, although the boundaries of the intervals only differ by the exponent $1$, let's have a closer look to both components of the formula:\\
(1) $x$ and (2) $\ln{x}$.

@@

\subsubsection*{The function $\frac{x}{\ln{x}}$}
The function $\frac{x}{\ln{x}}$ consists of the function $x$ as the numerator and the function $\ln{x}$ in the denominator, which, in comparison to $x$, increases very slowly. Compared to the number $x$ itself, the number of primes less or equal to $x$ is small.But still, $\frac{x}{\ln{x}}$ is an increasing function as you can see in figure \ref{xundlnxundxdurchlnx} on page \pageref{xundlnxundxdurchlnx}.

>>>>

To understand, why the number of primes is growing so rapidly, although the boundaries
of the intervals only differ by the exponent $1$, let's have a closer look to both
components of the right side of the formula: $x$ and $\ln{x}$.

@@

\subsubsection*{The function $\frac{x}{\ln{x}}$}
The function $\frac{x}{\ln{x}}$ consists of the function $x$ as the numerator and the function $\ln{x}$ in the denominator, which, in comparison to $x$, increases very slowly. Compared to the number $x$ itself, the number of primes less or equal to $x$ is small. But still, $\frac{x}{\ln{x}}$ is an increasing function as you can see in figure \ref{xundlnxundxdurchlnx} on page \pageref{xundlnxundxdurchlnx}.
